Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labelled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labelled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labelled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labelled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labelled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

Table~\ref{tab:experimental-results}  compares the  garbage collection
statisctics for  RGC and LGC  with reachability for  copying closures.
We report the  number of GC events, average number  of cells reclaimed
per  GC, average  number  of cells  touched per  GC,  total number  of
dragged cells  and total time  to perform  all collections.  It  is no
surprise  that  number  of  cells  reclaimed per  GC  is  higher  (and
consequently the  number of  GCs is lower)  for LGC  based collection.
LGC also reduces  the number of dragged  cells significantly. However,
the cost for LGC is that per  GC execution time is more for LGC, which
results in higher overall execution  time for garbage collection, even
with reduced  number of collections.   However, LGC run time  is still
competitive (slowdown within 5X of RGC  for worst case), and is better
for   3    benchmarks   (best    case   2X   speedup).     Note   that
\verb@gc_bench@~\cite{gc_bench}   is  a   synthestic  benchmark   that
allocates complete binary  trees of various sizes.  These binary trees
are  never used  by the  program. As  a result,  the benchmark  highly
favours LGC.   We have included  this benchmark for  completeness only
and the numbers for it  are not considered representative of practical
programs.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by our  collector.  In  case of LGC  for eager
languages~\ref{asati14lgc} the gap was  very narrow and almost touched
the actual  liveness curve.  In  case of  lazy languages, this  gap is
largely   due  to   approximate  reachability   based  collection   of
closures. To implement liveness based  collection of closures, we need
to  record liveness  data  for  each closure  and  update  it at  each
evaluation  point.   These  space  and  run  time  overheads  make  it
infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
practical garbage collector.

@@

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
  \rho', S, H[\rho'(z) := (H(\rho'(x)),H(\rho'(y)))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \cred{\sigma, e_f}$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
\cred{\clazy\sigma, e_f}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
  \cred{\sigma', e_f'}$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \cred{\sigma, e_f}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
\end{document}

>>>>

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

@@

The strongly regular grammar is converted  into a set of NFAs, one for
each $\var{\Lanv{i}{x}}$.  The $\hookrightarrow$ simplification is now
done on the NFAs by  repeatedly introducing $\epsilon$ edges to bypass
pairs  of consecutive  edges labeled  \bcar\acar\ or  \bcdr\acdr\ and
constructing the $\epsilon$-closure. Since the $\epsilon$-closure step
may  create  more  adjacent  \bcar\acar\ and  \bcdr\acdr\  edges,  the
simplification is continued till a fixed point is reached, after which
the edges labeled \bcar \ and  \bcdr are deleted.  The details of the
algorithm, its correctness and termination proofs are given by Karkare
et.  al.~\cite{karkare07liveness,asati14lgc}.  The resulting automaton
has  edges labeled  with  \acar,  \acdr\ and  \clazy\  only. In  this
automaton, for every edge labeled \clazy, we check if the source node
of the edge has a valid path to  a final final state.  If yes, we mark
the mark  the source node as  final. Finally, we remove  all the edges
labeled  \clazy\  and  convert  the automaton  into  a  deterministic
automaton.    This   effectively  implements   the   $\hookrightarrow$
simplification rules for  \bcar, \bcdr, and \clazy\  to obtain forward
paths. While  checking for liveness  during GC, a  forward access-path
only if it can reach a final state.
Figure~\ref{fig:example-automata}(a)  shows the  NFA that  is obtained
from the grammar for \var{\Df{\length}{1}}.  The final DFA is shown in
Figure~\ref{fig:example-automata}(b).  This expectedly says that given
a  demand $\epsilon$  on  \length,  the liveness  of  its argument  is
$\acdr^{*}$  (the  spine  of   the  list  is  traversed).   Similarly,
Figure~\ref{fig:example-automata}(c)     shows     the     NFA     for
\var{\Lanv{9}{\pa}}.          The         DFA         shown         in
Figure~\ref{fig:example-automata}(d)  does  not   accept  any  forward
paths.   This  reflects  the  lazy  nature  of  our  language.   Since
\length\ does not  evaluate the member elements of  the argument list,
the  closure for  \pa\ is  never evaluated  and is  reclaimed whenever
liveness based GC is triggered beyond $\pi_9$.

@@

Table~\ref{tab:experimental-results}  compares the  garbage collection
statisctics for  RGC and LGC  with reachability for  copying closures.
We report the  number of GC events, average number  of cells reclaimed
per  GC, average  number  of cells  touched per  GC,  total number  of
dragged cells  and total time  to perform  all collections.  It  is no
surprise  that  number  of  cells  reclaimed per  GC  is  higher  (and
consequently the  number of  GCs is lower)  for LGC  based collection.
LGC also reduces  the number of dragged  cells significantly. However,
the cost for LGC is that per  GC execution time is more for LGC, which
results in higher overall execution  time for garbage collection, even
with reduced  number of collections.   However, LGC run time  is still
competitive (slowdown within 5X of RGC  for worst case), and is better
for   3    benchmarks   (best    case   2X   speedup).     Note   that
\verb@gc_bench@~\cite{gc_bench}   is  a   synthestic  benchmark   that
allocates complete binary trees of  various sizes.  These binary trees
are  never used  by the  program. As  a result,  the benchmark  highly
favours LGC.   We have included  this benchmark for  completeness only
and the numbers for it  are not considered representative of practical
programs. All  our benchmarks were executed  on a desktop with  64 bit
Ubuntu 14.04, 8 core Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz, L2 cache
of 8192 KB and 15922 MiB (16 GB) RAM.

@@

One area  of concern is the  huge gap between the  actual liveness and
the liveness  percieved by our  collector.  In  case of LGC  for eager
languages~\ref{asati14lgc} the gap was  very narrow and almost touched
the actual  liveness curve.  In  case of  lazy languages, this  gap is
largely   due  to   approximate  reachability   based  collection   of
closures. To implement liveness based  collection of closures, we need
to  record liveness  data  for  each closure  and  update  it at  each
evaluation  point.   These  space  and  run  time  overheads  make  it
infeasible  to  incorporate  liveness  based  GC  for  closures  in  a
practical garbage collector.
>>>>>>> 86509285950f09b0dc95515465513023c4b6fa40

@@

\begin{figure*}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \cred{\sigma, e_f}$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
\cred{\clazy\sigma, e_f}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
  \cred{\sigma', e_f'}$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \cred{\sigma, e_f}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics for the language.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}
\end{document}
